Practical Optional Types for Clojure

### 2015-07-11 formal-model.tex
final POPL submission

A lookup of a key that is neither present or absent is checked by
T-GetHMapPartialDefault.
$$
\abs{\makelocal{u}}{\Lunch}{\getexp{\makelocal{u}}{\makekw{bocce}}}
$$
The result type is \Top, and the return object is \path{\keype{\makekw{bocce}}}{\makelocal{u}}.

@@

Extending a map with T-Assoc preserves its completeness.
$$
\abs{\makelocal{b}}{\Breakfast}{\assocexp{\makelocal{b}}{\makekw{au}}{\makestr{beans}}}
$$
The result type is
$
\HMapc{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}
        ,{\mandatoryentrynoarrow{\Valkw{au}}{\String}}}}
$.

@@

%The objects in the T-Get rules are more complicated than those in T-Local---the 
%next section discusses this in detail.
Finally T-AssocHMap extends an HMap with a mandatory entry while preserving completeness
and absent entries, and enforcing ${\k{}} \not\in {\absent{}}$ to prevent badly
formed types.

@@

The proof system for occurrence typing uses standard propositional logic,
except for where nested information is combined. This is
handled by the L-Update rule (\figref{main:figure:lupdate})---it says
under \propenv{}, if object \path{\pathelemp{}}{\x{}} is of type \t{}, and 
some extension
\path{\pathelem{}}{\path{\pathelemp{}}{\x{}}}
is of possibly-negative type \propisnotmeta{}, then
{\update{\t{}}{\propisnotmeta{}}{\pathelem{}}}
is the new type of \path{\pathelemp{}}{\x{}} under \propenv{}.

@@

In \egref{example:desserts-on-class},
$
\inpropenv{\propenv{}}{\isprop{\update{\Order}{\Long}{[{\classpe{}}, {\keype{\makekw{desserts}}}]}}{\makelocal{o}}}
$
updates the argument in the {\Long} method.
This recurs and extra time for each meal to handle the {\classpe{}}
path element.
%$
%{\update{\Number}{\Long}{[{\classpe{}}]}}}
%$
%
%$
%{\update{\Int}{\Long}{[{\classpe{}}]}}}
%$
%
%

>>>>

A lookup of a key that is neither present or absent is checked by
T-GetHMapPartialDefault.
$$
\abs{\makelocal{u}}{\Lunch}{\getexp{\makelocal{u}}{\makekw{bocce}}}
$$
The result type is \Top, and the return object is \path{\keype{\makekw{bocce}}}{\makelocal{u}}.
Notice the propositions attached to each \getliteral{} rule are trivial---propositions
are erased once they enter a HMap type.

@@

Extending a map with T-AssocHMap preserves its completeness.
$$
\abs{\makelocal{b}}{\Breakfast}{\assocexp{\makelocal{b}}{\makekw{au}}{\makestr{beans}}}
$$
The result type is
$
\HMapc{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}
        ,{\mandatoryentrynoarrow{\Valkw{au}}{\String}}}}
$,
which is complete, like its input.
T-AssocHMap also enforces ${\k{}} \not\in {\absent{}}$ to prevent badly formed types.

@@

%The objects in the T-Get rules are more complicated than those in T-Local---the 
%next section discusses this in detail.
%Finally T-AssocHMap extends an HMap with a mandatory entry while preserving completeness
%and absent entries, and enforcing ${\k{}} \not\in {\absent{}}$ to prevent badly
%formed types.

@@

The proof system for occurrence typing uses standard propositional logic,
except for where nested information is combined. This is
handled by the L-Update rule (\figref{main:figure:lupdate})---it says
under \propenv{}, if object \path{\pathelemp{}}{\x{}} is of type \t{}, and 
an extension
\path{\pathelem{}}{\path{\pathelemp{}}{\x{}}}
is of possibly-negative type \propisnotmeta{}, then
{\update{\t{}}{\propisnotmeta{}}{\pathelem{}}}
is \path{\pathelemp{}}{\x{}}'s type under \propenv{}.

@@

In \egref{example:desserts-on-class},
$
\inpropenv{\propenv{}}{\isprop{\update{\Order}{\Long}{[{\classpe{}}, {\keype{\makekw{desserts}}}]}}{\makelocal{o}}}
$
updates the argument in the {\Long} method.
This recurs twice for each meal to handle the {\classpe{}}
path element.

We describe the other \updateliteral{} cases.
The first \classpe{} case updates
to \class{} if \classconst{} returns \Value{\class{}}.
The second \keype{\k{}} case detects contradictions in absent
keys. % not overlapping with \Nil{}.
The third \keype{\k{}} case updates unknown entries to be mapped to \Nil{} or absent.
The fourth \keype{\k{}} case updates unknown entries to be \emph{present}
when they do no overlap with \Nil{}.

%$
%{\update{\Number}{\Long}{[{\classpe{}}]}}}
%$
%
%$
%{\update{\Int}{\Long}{[{\classpe{}}]}}}
%$
%
%
